---
title: "Resultaten & Conclusion"
author: "Jorick Baron"
date: '2022-10-03'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(stringr)
library(knitr)
library(kableExtra)
library(e1071) 
library(foreign)
```

\pagebreak

# Results

## Codebook

To aid in the understanding of the data a codebook explaining every variable in the data is displayed bellow.

```{r codebook, echo=FALSE}
codebook <- read.delim("Data/codebook.csv", sep = ",")
kable(codebook, caption = "Codebook", align = "lcccr", booktabs = T) %>%
  kable_styling(latex_options = c("scale_down"))
```

## Loading The Data

For this project the data from <https://www.kaggle.com/johnjdavisiv/urinary-biomarkers-for-pancreatic-cancer> was used. Upon loading the data the first few data entries where checked using the head function, this revealed many NAs.

```{r head, echo=FALSE}
data <- read.table(file="Data/source/Debernardi_et_al_2020_data.csv", sep = ",",
                   header = T, na.strings = "")
head(data)
```

The ones from the stage column can be explained because only patients with cancer can have a stage. And those from benign_sample_diagnosis are there because there has to be a non-cancerous diagnosis. Reading in the source the NAs in the colums plasma_CA19_9 and REG1A are there because they were not measured in every sample, thus they contain NAs. From this the conclusion is that the data is intact and further processing can be applied.

## Exploring & Analysing

When exploring the data for processing a boxplot to investigate the distribution of every numeric variable was created.

```{r boxplot, echo=FALSE, fig.height=5, fig.width=10, fig.cap="A boxplot displaying the distribution of all numeric values in the data", warning=FALSE}
p1<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = age))+
  ylab("age in years") + 
  xlab(NULL)

p2<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = plasma_CA19_9))+
  ylab("plasma_CA19_9 in U/ml") + 
  xlab(NULL)+
  ylim(0,500)

p3<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = creatinine))+
  ylab("creatinine in mg/ml") + 
  xlab(NULL)

p4<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = LYVE1))+
  ylab("LYVE1 in ng/ml") + 
  xlab(NULL)+
  ylim(0,17)

p5<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = REG1B))+
  ylab("REG1B in ng/ml") + 
  xlab(NULL)+
  ylim(0,600)

p6<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = TFF1))+
  ylab("TFF1 in ng/ml") + 
  xlab(NULL)+
  ylim(0,5000)

p7<- ggplot(data=data)+
  geom_boxplot(mapping = aes(x = "",
                            y = REG1A))+
  ylab("REG1A in ng/ml") + 
  xlab(NULL) +
  ylim(0,5000)

grid.arrange(p1, p2, p3, p4, p5, p6, p7, nrow = 2)
```

\pagebreak

In this figure it became clear that the distributions of most variables seem highly skewed, to investigate this another boxplot was made to see if this was due to the influence of the label

```{r boxplot grouped, fig.height=10, fig.width=20, message=FALSE, warning=FALSE, fig.cap="boxplots with added dimension (diagnosis)", echo=FALSE}
data$diagnosis <- factor(data$diagnosis)

gp1 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = age,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("age in years")

gp2 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = plasma_CA19_9,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("plasma_CA19_9 in U/ml")+
  ylim(0,500)

gp3 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = creatinine,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("creatinine in mg/ml")

gp4 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = LYVE1,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("LIVE1 in ng/ml")+
  ylim(0,17)

gp5 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = REG1B,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("REG1B in ng/ml")+
  ylim(0,600)

gp6 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = TFF1,
                            group=diagnosis,
                            fill=diagnosis))+
  scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("TFF1 in ng/ml")+
  ylim(0,5000)

gp7 <- ggplot(data = data)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = REG1A,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("REG1A in ng/ml")+
  ylim(0,5000)

grid.arrange(gp1, gp2, gp3, gp4, gp5, gp6, gp7, nrow = 4)
```

\pagebreak

It seems something else is at play to cause this perceived skewness. A statistical test using the skewness function in R can be used to test if the data is skewed and in what direction.

```{r skew, echo=FALSE}
paste("skewness of age:", skewness(data$age))
paste("skewness of plasma_CA19_9:", skewness(data$plasma_CA19_9, na.rm = T))
paste("skewness of creatinine:", skewness(data$creatinine))
paste("skewness of LYVE1:", skewness(data$LYVE1))
paste("skewness of REG1B:", skewness(data$REG1B))
paste("skewness of TFF1:", skewness(data$TFF1))
paste("skewness of REG1A:", skewness(data$REG1A, na.rm = T))
```

This test proves that all but age is significantly skewed, since everything greater than a 1 or smaller than -1 is significant. To combat this skewness the data can be log transformed, scaling it logarithmically makes sense when the data can not be negative and seems to have expectational differences.

```{r trans boxplots, echo=FALSE, fig.height=10, fig.width=15, message=FALSE, warning=FALSE, fig.cap="boxplot of log transformed data"}
trans <- as.data.frame(log(data$plasma_CA19_9))
names(trans) <- "plasma_CA19_9"
trans$creatinine <- log(data$creatinine)
trans$LYVE1 <- log(data$LYVE1)
trans$REG1B <- log(data$REG1B)
trans$TFF1 <- log(data$TFF1)
trans$REG1A <- log(data$REG1A + 1)
trans$diagnosis <- data$diagnosis

tgp1 <- ggplot(data = trans)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = plasma_CA19_9,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("plasma_CA19_9 in U/ml (log transformed)")

tgp2 <- ggplot(data = trans)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = creatinine,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("creatinine in mg/ml (log transformed)")

tgp3 <- ggplot(data = trans)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = LYVE1,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("LIVE1 in ng/ml (log transformed)")

tgp4 <- ggplot(data = trans)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = REG1B,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("REG1B in ng/ml (log transformed)")

tgp5 <- ggplot(data = trans)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = TFF1,
                            group=diagnosis,
                            fill=diagnosis))+
  scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("TFF1 in ng/ml (log transformed)")

tgp6 <- ggplot(data = trans)+
  geom_boxplot(mapping = aes(x = diagnosis,
                            y = REG1A,
                            group=diagnosis,
                            fill=diagnosis))+
    scale_x_discrete(labels=c("control",
                            "benign hepatobiliary disease",
                            "pancreatic cancer"))+
  xlab("diagnosis")+
  ylab("REG1A in ng/ml (log transformed)")

grid.arrange(tgp1, tgp2, tgp3, tgp4, tgp5, tgp6, nrow = 3)
```

\pagebreak

looking at the data now it seems much less skewed however after a data transformation the normalcy should be tested this can be done using both a qq_plot and a shapiro.test.

```{r normalcy, fig.height=10, fig.width=20, message=FALSE, warning=FALSE, fig.cap="qqplots displaying normalcy", echo=FALSE}
shapiro.test(trans$plasma_CA19_9)
qq1 <- ggplot(trans, aes(sample = plasma_CA19_9, colour = diagnosis)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("plasma_CA19_9")

shapiro.test(trans$creatinine)
qq2 <- ggplot(trans, aes(sample = creatinine, colour = diagnosis)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("creatinine")

shapiro.test(trans$LYVE1)
qq3 <- ggplot(trans, aes(sample = LYVE1, colour = diagnosis)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("LYVE1")

shapiro.test(trans$REG1B)
qq4 <- ggplot(trans, aes(sample = REG1B, colour = diagnosis)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("REG1B")

shapiro.test(trans$TFF1)
qq5 <- ggplot(trans, aes(sample = TFF1, colour = diagnosis)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("TFF1")

shapiro.test(trans$REG1A)
qq6 <- ggplot(trans, aes(sample = REG1A, colour = diagnosis)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("REG1A")

grid.arrange(qq1, qq2, qq3, qq4, qq5, qq6, nrow = 2)
```

These variables are obviously not normally distributed due to none of them being above a p-value of 0.05, meaning that the null-hypothesis of the data being normaly distributed is rejected.

\pagebreak

The final analysis is to explore if any variables are correlated this can be displayed using a heat-map.

```{r heatmap, fig.height=10, fig.width=20, fig.cap="heatmap displaying correlation of values", echo=FALSE}
new_data <- cbind(data[3:6], trans[2:6])
matrix_data <- drop_na(new_data[,c(2, 5:9)])
cor_matrix <- cor(matrix_data)
heatmap(cor_matrix, scale = "column", col = heat.colors(5, rev = T), Colv = NA, Rowv = NA)
legend(x="right", legend=c("full","strong", "medium", "minimal", "none"),fill=heat.colors(5))
title(main = "Heatmap depicting correlations", line = 3, sub = "")
```

\pagebreak

REG1 A and B seem moderately correlated (`r cor_matrix[4,6]`), otherwise no real strong correlation is observed.

# Conclusion

## Final Results

The final data set has had all numeric values except age log-transformed to reduce the skew but this left the data unnormal and also columns: sample_id, patient_cohort, sample_origin, stage and benign_sample_diagnosis dropped because them not being suitable for machine learning, furthermore collum diagnosis has been replaced with has_cancer because the research is focused to detect cancer. And plasma_CA19_9 has been dropped because in this research we focus on data that can be gathered from a urine sample. The data should however be useful and relaiable for machine learning.
